---
title: "Explore Relationships and Ancestors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explore Relationships and Ancestors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(chariot)
library(tidyverse)
library(RColorBrewer)
```

```{r}
nb.cols <- 20
mycolors <- colorRampPalette(brewer.pal(8, "Paired"))(nb.cols)
```

There are 4 different types of relationships represented in 
the Concept Relationship table depending on whether they are 
hierarchical and if they define ancestry:  

1. Is Hierarchical & Does Not Define Ancestry  
1. Is Not Hierarchical & Defines Ancestry  
1. Is Hierarchical & Defines Ancestry   
1. Is Not Hierarchical & Does Not Define Ancestry  

These relationship types can be queried directly in the 
Relationship table.  

```sql
SELECT DISTINCT 
  is_hierarchical,
  defines_ancestry
FROM omop_vocabulary.relationship
```

```{r,message=FALSE,results='hide'}
relationship_types <- 
  queryAthena(sql_statement = 
                "
                SELECT DISTINCT 
                  is_hierarchical,
                  defines_ancestry
                FROM omop_vocabulary.relationship
                ")
```

The relationship types are presented as boolean with the 
following definitions from the [OMOP CDM v5.3.1 wiki](https://ohdsi.github.io/CommonDataModel/cdm531.html#RELATIONSHIP):  
**is_hierarchical**: Defines whether a relationship defines 
concepts into classes or hierarchies. Values are 1 for 
hierarchical relationship or 0 if not.  
**defines_ancestry**: Defines whether a hierarchical 
relationship contributes to the concept_ancestor table. 
These are subsets of the hierarchical relationships. 
Valid values are 1 or 0.	

```{r}
relationship_types
```

Though a relationship that defines ancestry is by definition 
required to also be deemed hierarchical, non-hierarchical 
relationships that define ancestry appear to exist in the 
Relationship table. 

```sql
SELECT DISTINCT
  c.vocabulary_id AS vocabulary_id_1,
  c.concept_class_id AS concept_class_id_1,
  c2.vocabulary_id AS vocabulary_id_2,
  c2.concept_class_id AS concept_class_id_2,  
  cr.relationship_id,
  cr.invalid_reason
FROM omop_vocabulary.relationship r 
INNER JOIN omop_vocabulary.concept_relationship cr 
ON cr.relationship_id = r.relationship_id 
INNER JOIN omop_vocabulary.concept c 
ON cr.concept_id_1 = c.concept_id 
INNER JOIN omop_vocabulary.concept c2 
ON cr.concept_id_2 = c2.concept_id 
WHERE 
  r.is_hierarchical = 0 
    AND r.defines_ancestry = 1 
    AND c.invalid_reason IS NULL 
    AND c2.invalid_reason IS NULL
```

```{r,results='hide'}
outliers <-
  queryAthena(sql_statement = 
              "
              SELECT DISTINCT
                c.vocabulary_id AS vocabulary_id_1,
                c.concept_class_id AS concept_class_id_1,
                c2.vocabulary_id AS vocabulary_id_2,
                c2.concept_class_id AS concept_class_id_2,  
                cr.relationship_id,
                cr.invalid_reason
              FROM omop_vocabulary.relationship r 
              INNER JOIN omop_vocabulary.concept_relationship cr 
              ON cr.relationship_id = r.relationship_id 
              INNER JOIN omop_vocabulary.concept c 
              ON cr.concept_id_1 = c.concept_id 
              INNER JOIN omop_vocabulary.concept c2 
              ON cr.concept_id_2 = c2.concept_id 
              WHERE 
                r.is_hierarchical = '0'
                  AND r.defines_ancestry = '1' 
                  AND c.invalid_reason IS NULL 
                  AND c2.invalid_reason IS NULL
              ")  %>%
    mutate(`Vocabulary to Vocabulary` = paste0(vocabulary_id_1, " to ", vocabulary_id_2))
```

The relationships indicate that these types of relationships only 
exist in the Drug Domain. 

```{r,echo=FALSE}
# Since fill is length > 8, a new set of colors must be made for 
# the length of the field that will be the fill 
nb.cols <- length(unique(outliers$`Vocabulary to Vocabulary`))
mycolors <- colorRampPalette(brewer.pal(8, "Set2"))(nb.cols)

# Plot
ggplot2::ggplot(
  data = outliers,
  mapping = aes(x = relationship_id,
                fill = `Vocabulary to Vocabulary`)) +
  ggplot2::geom_bar() + 
  ggplot2::scale_fill_manual(values = mycolors) +
  ggplot2::theme_minimal() + 
  ggplot2::labs(
    title = "Non-Hierarchical Relationships that Define Ancestry",
    x = "Relationship Id") +
  ggplot2::theme(
    axis.text.x = element_text(angle = 30, 
                               vjust = 1,
                               size = 6),
    legend.text = element_text(size = 6),
    legend.key.size = unit(.33, "cm")
    )
```


## Full Review  

For each of the 4 relationship types, any association with select 
controlled vocabularies SNOMED, LOINC, RxNorm, RxNorm Extension, 
HemOnc and ATC is derived. Relationships analyzed are filtered 
for those within each vocabulary system.  

### Methods  

Relationships for each relationship type and target vocabulary 
is queried. 

```sql
SELECT 
  c.vocabulary_id,
  c.concept_class_id AS concept_class_id_1,
  c2.concept_class_id AS concept_class_id_2,
  r.relationship_id,
  r.is_hierarchical,
  r.defines_ancestry,
  ca.min_levels_of_separation,
  ca.max_levels_of_separation
FROM patelm9.@staging_table t 
INNER JOIN omop_vocabulary.relationship r 
ON t.is_hierarchical = r.is_hierarchical 
    AND t.defines_ancestry = r.defines_ancestry 
INNER JOIN omop_vocabulary.concept_relationship cr 
ON cr.relationship_id = r.relationship_id 
INNER JOIN omop_vocabulary.concept c 
ON cr.concept_id_1 = c.concept_id  
INNER JOIN omop_vocabulary.concept c2 
ON cr.concept_id_2 = c2.concept_id 
INNER JOIN omop_vocabulary.concept_ancestor ca 
ON ca.ancestor_concept_id = cr.concept_id_1 
     AND ca.descendant_concept_id = cr.concept_id_2
WHERE 
  c.invalid_reason IS NULL 
    AND c2.invalid_reason IS NULL 
    AND c.vocabulary_id IN ('@target_vocabulary') 
    AND c2.vocabulary_id IN ('@target_vocabulary')
    AND cr.invalid_reason IS NULL
```

```{r,results='hide',cache=TRUE}
# Target vocabularies
target_vocabularies <- c("SNOMED", "LOINC", 
                         "RxNorm", "RxNorm Extension", 
                         "HemOnc", "ATC")

# Looping over each relationship type 
output <- list()
for (i in 1:nrow(relationship_types)) {
  relationship_type <- 
    relationship_types %>%
    filter(row_number() == i)
  
  conn <- connectAthena()
  staging_table <- pg13::write_staging_table(conn = conn, 
                                             schema = "patelm9",
                                             data = relationship_type,
                                             drop_existing = TRUE,
                                             drop_on_exit = FALSE)
  
  output[[i]] <- list()
  # Loop over each target vocabulary by relationship type
  for (j in seq_along(target_vocabularies)) {
  
    target_vocabulary <- target_vocabularies[j]
    
    sql_statement <-
      SqlRender::render(
        "
        SELECT 
          c.vocabulary_id,
          c.concept_class_id AS concept_class_id_1,
          c2.concept_class_id AS concept_class_id_2,
          r.relationship_id,
          r.is_hierarchical,
          r.defines_ancestry,
          ca.min_levels_of_separation,
          ca.max_levels_of_separation
        FROM patelm9.@staging_table t 
        INNER JOIN omop_vocabulary.relationship r 
        ON t.is_hierarchical = r.is_hierarchical 
            AND t.defines_ancestry = r.defines_ancestry 
        INNER JOIN omop_vocabulary.concept_relationship cr 
        ON cr.relationship_id = r.relationship_id 
        INNER JOIN omop_vocabulary.concept c 
        ON cr.concept_id_1 = c.concept_id  
        INNER JOIN omop_vocabulary.concept c2 
        ON cr.concept_id_2 = c2.concept_id 
        INNER JOIN omop_vocabulary.concept_ancestor ca 
        ON ca.ancestor_concept_id = cr.concept_id_1 
             AND ca.descendant_concept_id = cr.concept_id_2
        WHERE 
          c.invalid_reason IS NULL 
            AND c2.invalid_reason IS NULL 
            AND c.vocabulary_id IN ('@target_vocabulary') 
            AND c2.vocabulary_id IN ('@target_vocabulary')
            AND cr.invalid_reason IS NULL
        ",
        staging_table = staging_table,
        target_vocabulary = target_vocabulary
      )
    
    output[[i]][[j]] <-
      queryAthena(sql_statement = sql_statement)
    names(output[[i]])[j] <- target_vocabulary
  
  }
  
  pg13::drop_all_staging_tables(conn = conn, 
                                schema = "patelm9")
  dcAthena(conn = conn)
}

# Add names to output
relationship_type_labels <- 
  c("Is Hierarchical & Does Not Define Ancestry",
    "Is Not Hierarchical & Defines Ancestry",
    "Is Hierarchical & Defines Ancestry", 
    "Is Not Hierarchical & Does Not Define Ancestry")
names(output) <- relationship_type_labels
```

The relationships are counted by relationship type.  

```{r}
# Bind outputs by relationship type
output2_a <-
  output %>%
  map(bind_rows) %>%
  # Rename target fields for aesthetics in plots
  map(function(x) x %>% rename(Vocabulary = vocabulary_id)) %>%
  # Add relationship_id count
  map(group_by,relationship_id) %>%
  map(mutate, count = length(relationship_id)) %>%
  map(ungroup) %>%
  # Convert controlled vocabularies to factors
  map(mutate_at, vars(Vocabulary), ~ factor(., 
                                       levels = c("SNOMED",
                                                  "LOINC",
                                                  "RxNorm",
                                                  "RxNorm Extension",
                                                  "ATC", 
                                                  "HemOnc")))
```

The distribution of `max_levels_of_separation` and 
`min_levels_of_separation` is also assessed by relationship 
type.  

```{r}
output2_b <-
  output2_a %>%
  map(function(x) 
    x %>% 
        pivot_longer(cols = c(min_levels_of_separation, 
                              max_levels_of_separation),
                     names_to = "Type of Separation",
                     values_to = "level_of_separation") %>%
      mutate_at(vars(`Type of Separation`),
              function(y) case_when(y %in% c("max_levels_of_separation") ~ "Max",
                                    TRUE ~ "Min")))
```


```{r,echo=FALSE}
plot_counts <- 
  function(data,
           title) {
    ggplot2::ggplot(data = data,
                aes(x = reorder(relationship_id, -count),
                    fill = Vocabulary)) +
  ggplot2::geom_bar() + 
  ggplot2::scale_fill_brewer(palette = "Paired") +
  ggplot2::labs(title = title,
                x = "Relationship Id") + 
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = element_text(angle = 30, 
                                            vjust = 1,
                                            size = 6),
                 legend.text = element_text(size = 6),
                 legend.key.size = unit(.33, "cm"))
  }
```

```{r,echo=FALSE}
plot_levels <-
  function(data,
           title) {
  ggplot2::ggplot(data = data,
                  aes(x = level_of_separation, 
                      fill = `Type of Separation`)) +
    ggplot2::geom_bar(position = "dodge") + 
    ggplot2::scale_fill_brewer(palette = "Dark2") +
    ggplot2::labs(title = title,
                  x = "Level of Separation")
  }
```

### Results  
#### Hierarchical & Does Not Define Ancestry 

There are zero relationships that are hierarchical, but do 
not define ancestry in the select controlled vocabularies.  

#### Not Hierachical & Define Ancestry  

Though already reviewed earlier, relationships that are not 
hierarchical and define ancestry is re-reviewed for the 
controlled vocabulary subset of RxNorm, RxNorm Extension, ATC, 
and HemOnc. 

```{r,echo=FALSE}
plot_counts(output2_a$`Is Not Hierarchical & Defines Ancestry`,
            title = "Relationship Id Distribution")
```

```{r,echo=FALSE}
plot_levels(output2_b$`Is Not Hierarchical & Defines Ancestry`,
            title = "Levels of Separation Distribution")
```

#### Not Hierachical & Define Ancestry 

```{r,echo=FALSE}
plot_counts(data = output2_a$`Is Hierarchical & Defines Ancestry`,
            title = "Relationship Id Distribution")
```

```{r,echo=FALSE}
plot_levels(
  data = output2_b$`Is Hierarchical & Defines Ancestry`,
  title = "Levels of Separation Distribution"
  )
```

#### Not Hierarchical and Do Not Define Ancestry  

```{r,echo=FALSE}
plot_counts(
    data = output2_a$`Is Not Hierarchical & Does Not Define Ancestry`,
    title = "Relationship Id Distribution"
  )
```

```{r,echo=FALSE}
plot_levels(
  data = output2_b$`Is Not Hierarchical & Does Not Define Ancestry`,
  title = "Levels of Separation Distribution"
  )
```


