---
title: "Harmonizing HemOnc Concepts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Harmonizing HemOnc Concepts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(chariot)
library(tidyverse)
```

Harmonizing HemOnc concepts entails getting a combination of Regimens and 
Components, reducing the combination down to the Component level, and finding 
Regimens and/or Regimen and Component combinations that map to the reduced list. 

For example, the Regimen FOLFOX along with Irinotecan would harmonize to 
FOLFIRINOX. 

```{r}
FOLFOX <- get_concept(concept_id = 35806596)
Irinotecan <- get_concept(concept_id = 35803130)
```

To check for this, the FOLFOX plus Irinotecan combination are reduced to Components. 

```{r}
new_components <- 
  ho_reduce_to_components(FOLFOX,
                        Irinotecan)
```

```{r}
new_components
```

The components can then be submitted to generate a potential regimen match. 

```{r}
new_regimen <- 
  ho_lookup_regimen(new_components$concept_id)
```

```{r}
new_regimen
```


In this case, various combinations existed. However, what if a combination does 
not cleanly map to a HemOnc Regimen? For example, sometimes there is an add-on 
drug to an established Regimen for experimental reasons.  

For demonstration purposes, suppose the new Components reduced down to 
Fluorouracil, Folinic acid, Oxaliplatin, and Irinotecan as well as Trastuzumab. 

```{r}
Trastuzumab <- get_concept(concept_id = 35803361)
Trastuzumab
```


```{r}
new_component_ids2 <- c(new_components$concept_id, Trastuzumab@concept_id)
new_component_ids2
```


```{r}
new_regimen2 <- ho_lookup_regimen(new_component_ids2)
```

In this case, a regimen was not found as indicated by a zero row result. 
```{r}
new_regimen2
```

To find possible Regimen and Component combinations instead, subsets of the 
Components can be submitted to find matches using `ho_grep_regimens()` with 
starting `component_count` 1 value less than the total Component count. In 
this case, it is 4. 

The Concept classes are retrieved for each of the Components. 

```{r}
new_component_objs2 <- lapply(new_component_ids2,
                              get_concept)
names(new_component_objs2) <- lapply(new_component_objs2,
                                     function(x) slot(x, "concept_name"))
new_component_objs2
```

All the possible Component permutations are made by removing one of the Components. 

```{r}
new_component_permutations <- list()
for (i in seq_along(new_component_objs2)) {
  new_component_permutations[[i]] <-
  new_component_objs2[-i]
  names(new_component_permutations)[i] <- new_component_objs2[[i]]@concept_name
}
```

The resulting input is of the same length as the source Components (5) and each 
value in the list is composed of 4 (5 minus 1) Components. The names of the 
removed Component in a particular permutation is also saved as the name. 

The permutations are mapped to Regimens.

```{r}
new_regimens2 <- list()
for (i in seq_along(new_component_permutations)) {
  
  new_regimens2[[i]] <-
  ho_lookup_regimen(new_component_permutations[[i]])
}
names(new_regimens2) <- names(new_component_permutations)
new_regimens2 <-
  new_regimens2 %>%
  purrr::keep(~ nrow(.) > 0)
new_regimens2
```

Filtering for only the permutations that returned greater than zero rows, 
there are 3 potential regimens that a subset of 4 Components map to:  

```{r}
new_regimens2 %>% 
  bind_rows(.id = "add_on_component_name") %>%
  select(add_on_component_name, regimen_concept_id, regimen_concept_name) %>%
  distinct()
```

Therefore, the possible combinations are either FOLFIRINOX plus Trastuzumab, 
mFOLFIRINOX plus Trastuzumab, or FOLFIRINOX/modified FOLFIRINOX plus /- Chemoradiation 
plus Trastuzumab.  

It is possible that the permutations do not return any rows in this scenario. If 
this is the case, it is likely more feasible to manually look up the mapping at 
\url{athena.ohdsi.org} than to programmatically find matches. 
