---
title: "Pivot Concept Relationships"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pivot Concept Relationships}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,message=FALSE,warning=FALSE,results='hide'}
library(chariot)
library(tidyverse)
```

```{r,results='hide', message=FALSE}
conn <- connectAthena()
```

Pivoting the Concept Relationship table can provide a rich array representing 
the various relationships an OMOP Concept may have. 

To exemplify, a test dataset of RxNorm and RxNorm Extension concepts are 
retrieved. 

```{r}
test_data <- 
  get_test_drug_concepts(conn = conn)
```

```{r}
test_data
```

The relationships of these concepts are derived. 

```{r}
test_data_relationships <- lookup_relationships(test_data$concept_id,
                                                conn = conn)
```

```{r}
test_data_relationships
```

To properly pivot the resultset, the concepts on both sides require transformation 
into a single cell per concept. This is achieved by converting concepts into the 
`strip` format. 

```{r}
test_data_relationships2 <-
  test_data_relationships %>%
  merge_strip(into = "concept_1", 
              suffix = "_1") %>%
  merge_strip(into = "concept_2",
              suffix = "_2")
```

```{r}
test_data_relationships2
```

The dataset can now be pivoted by the relationship, hinging on `concept_1`, which 
are the concepts that were originally inputted. 

```{r}
output <-
  test_data_relationships2 %>%
  pivot_wider(id_cols = concept_1,
              names_from = relationship_id, 
              values_from = concept_2)
output
```

Since at times a `concept_1` may have a specific relationship to more than 1 
`concept_2`, a list-cols are returned in the output. To create a more readable 
output, an aggregate function on `concept_2` must be supplied to convert multiple 
`concept_2` values into a string of length 1. 

```{r}
output <-
  test_data_relationships2 %>%
  pivot_wider(id_cols = concept_1,
              names_from = relationship_id, 
              values_from = concept_2,
              # Add aggregate function
              values_fn = list(concept_2 = ~ paste(unique(.), collapse = "|"))
              )
output
```

The resulting dataset contains a row for each original concept provided with 
each relationship and concept/s it has the relationship to in a pipe-separated 
string. 

```{r,include=FALSE}
dcAthena(conn = conn)
```



