---
title: "Pivot Concept Ancestors and Descendants"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pivot Concept Ancestors and Descendants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,message=FALSE,warning=FALSE,results='hide'}
library(chariot)
library(tidyverse)
```

```{r,results='hide', message=FALSE}
conn <- connectAthena()
```

Pivoting the Concept Ancestor table can provide a rich array representing 
the class hierarchy an OMOP Concept may have. 

To exemplify, a test Drug concept is retrieved. Using more than 1 concept can 
lead to messy results since the classifications between randomly acquired concepts 
can be vastly different. 

```{r}
test_data <- 
  get_test_drug_concepts(conn = conn,
                         limit = 1)
```

```{r}
test_data
```

The ancestors of this concept are derived. 

```{r}
test_data_ancestors <-
  join_for_ancestors(data = test_data,
                     descendant_id_column = "concept_id",
                     conn = conn)
```

```{r}
test_data_ancestors
```

The descendants are also derived. 

```{r}
test_data_descendants <-
  join_for_descendants(
    data = test_data, 
    ancestor_id_column = "concept_id",
    conn = conn
  )
```

```{r}
test_data_descendants
```

The construct a full concept lineage, the ancestors and descendants are harmonized. 
Harmonization is achieved by:  

1. Renaming the `ancestor_` and `descendant_` concepts  
to `relative_` while maintaining the original identity in a separate column  
1. Arranging the ancestors in descending order by levels of separation and the 
descendants in ascending order   
1. Combining the two datasets to create the complete lineage from the topmost 
ancestor to the bottommost descendant  

```{r}
test_data_ancestors2 <- 
  test_data_ancestors %>%
  rename_all(str_replace_all, "ancestor_", "relative_") %>% 
  mutate(relative_type = "ancestor") %>%
  arrange(desc(min_levels_of_separation))

test_data_descendants2 <-
  test_data_descendants %>%
  rename_all(str_replace_all, "descendant_", "relative_") %>% 
  mutate(relative_type = "descendant") %>%
  arrange(min_levels_of_separation)

test_data_relatives <-
  bind_rows(test_data_ancestors2,
            test_data_descendants2)

test_data_relatives
```

The preserve the ordering of the lineage, a `levels_of_separation` field of 
the factor datatype is created by combining a marker between `relative_type` and 
the `min_levels_of_separation`.

```{r}
test_data_relatives2 <-
  test_data_relatives %>%
  unite(col = levels_of_separation, 
        relative_type,
        min_levels_of_separation) %>%
  mutate(levels_of_separation = factor(levels_of_separation,
                                       levels = c("ancestor_3", "ancestor_2",
                                                  "ancestor_1", "ancestor_0", 
                                                  "descendant_0", "descendant_1")))
test_data_relatives2$levels_of_separation
```


To properly pivot the resultset, the source concept and relative concepts 
require transformation into a single cell per concept. This is achieved by 
converting concepts into the `strip` format. 

```{r}
test_data_relatives3 <-
  test_data_relatives2 %>%
  merge_strip(into = "concept") %>%
  merge_strip(into = "relative", 
              prefix = "relative_")
```

```{r}
test_data_relatives3
```

The dataset can now be pivoted by the level of separation, hinging on the source 
`concept`.

```{r}
output <-
  test_data_relatives3 %>%
  pivot_wider(id_col = concept,
              names_from = levels_of_separation, 
              values_from = relative)
output
```

Since `concept` has more than 1 `relative` at any given `levels_of_separation`,
list-cols are returned in the output. To create a more readable  output, an 
aggregate function on `relative` must be supplied to convert multiple `relative` 
values into a string of length 1. 

```{r}
output <-
  test_data_relatives3 %>%
  pivot_wider(id_col = concept,
              names_from = levels_of_separation, 
              values_from = relative,
              #Add aggregate 
              values_fn = list(relative = ~ paste(unique(.), collapse = "|")))
output
```


The resulting dataset contains a single row for the source concept provided with 
each relative cast across in the order from topmost ancestor to bottommost 
descendant aggregated into a pipe-separated string. 

```{r,include=FALSE}
dcAthena(conn = conn)
```



