---
title: "Taxonomy and Relationships"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Taxonomy and Relationships}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  results = 'hide',
  cache   = TRUE,
  comment = "#>"
)

print_dt <-
  function(x,
           rownames = FALSE,
           filter = "top",
           scrollX = TRUE) {
    
    x <-
      x %>% 
      dplyr::mutate_all(factor)
    
    DT::datatable(
      x,
      rownames = rownames,
      filter = filter,
      extensions = "Buttons",
      options = 
        list(dom = "Blfrtip", 
             buttons = c("copy", "csv","excel", "pdf", "print"), 
             lengthMenu = list(c(10, 25, 50, -1), c(10, 25, 50, "All")),
             scrollX = scrollX
            )
      )
  }
```

```{r setup}
library(chariot)
```

3 types of relationships can be derived from the OMOP 
Vocabulary tables:  

1. [Taxonomy](#taxonomy): hierarchical parent-to-child relationships represented 
by _Is a_ or _Subsumes_ predicates.  
2. [Mapping](#mapping): _Maps to_ or _Mapped from_ relationships between 
OMOP concepts that either a.) represent a 
mapping from a non-standard vocabulary to a standard controlled vocabulary 
such as NDC to RxNorm or b.) represent a mapping from a deprecated 
concept to its most current counterpart.  
3. [Lateral](#lateral-relationships): non-hierarchical relationships between OMOP concepts 
represented by all other relationships. 

Here, I am using "[V] [S] 4115276 Non-small cell lung cancer [SNOMED 254637007] [Condition] [Clinical Finding]" 
as an example.  


# Taxonomy  

## CONCEPT_ANCESTOR Table  

The ancestry of 4115276 is derived and ordered in a top-to-bottom 
format. The CONCEPT_ANCESTOR table provides 2 types of paths: 
the `min_levels_of_separation` and `max_levels_of_separation` 
that represents the minimum and maximum distances between 
the concepts within the hierarchy, respectively. Here, the 
minimum is used.  


```sql
-- The ancestors are ordered from top-to-bottom
SELECT c.concept_name, ca.min_levels_of_separation 
FROM omop_vocabulary.concept_ancestor ca 
LEFT JOIN omop_vocabulary.concept c 
  ON c.concept_id = ca.ancestor_concept_id 
WHERE 
  ca.descendant_concept_id = 4115276 
    AND c.invalid_reason IS NULL 
ORDER BY min_levels_of_separation DESC
UNION
-- The descendants ordered inverse to maintain top-to-bottom format
SELECT c.concept_name, ca.min_levels_of_separation 
FROM omop_vocabulary.concept_ancestor ca 
LEFT JOIN omop_vocabulary.concept c 
ON c.concept_id = ca.descendant_concept_id 
WHERE ca.ancestor_concept_id = 4115276 
AND c.invalid_reason IS NULL 
ORDER BY min_levels_of_separation;
```


```{r}
taxonomy <- 
queryAthena(
  "
  WITH a AS (
    SELECT c.concept_name, ca.min_levels_of_separation 
    FROM omop_vocabulary.concept_ancestor ca 
    LEFT JOIN omop_vocabulary.concept c 
      ON c.concept_id = ca.ancestor_concept_id 
    WHERE 
      ca.descendant_concept_id = 4115276 
        AND c.invalid_reason IS NULL 
    ORDER BY min_levels_of_separation DESC
  ),
  d AS (
   SELECT c.concept_name, ca.min_levels_of_separation 
    FROM omop_vocabulary.concept_ancestor ca 
    LEFT JOIN omop_vocabulary.concept c 
    ON c.concept_id = ca.descendant_concept_id 
    WHERE ca.ancestor_concept_id = 4115276 
    AND c.invalid_reason IS NULL 
    ORDER BY min_levels_of_separation
  )
  
  SELECT * 
  FROM a 
  UNION 
  SELECT * 
  FROM d
  ",
  verbose = FALSE,
  render_sql = FALSE
)
```

```{r,echo=FALSE,results='markup'}
print_dt(taxonomy)
```



## CONCEPT_RELATIONSHIP Table  

The hierarchy seen above can also be derived from the CONCEPT_RELATIONSHIP 
table by filtering for the _Is a_ and _Subsumes_ relationships.  

```sql
SELECT relationship_id, c.* 
FROM omop_vocabulary.concept_relationship cr 
LEFT JOIN omop_vocabulary.concept c 
ON c.concept_id = cr.concept_id_2
WHERE 
  cr.concept_id_1 = 4115276 AND
  cr.invalid_reason IS NULL AND 
  relationship_id IN 
    ('Subsumes', 'Is a');
```

```{r}
taxonomy_relationship <-
queryAthena(
  "
  SELECT relationship_id, c.* 
  FROM omop_vocabulary.concept_relationship cr 
  LEFT JOIN omop_vocabulary.concept c 
  ON c.concept_id = cr.concept_id_2
  WHERE 
    cr.concept_id_1 = 4115276 AND
    cr.invalid_reason IS NULL AND 
    relationship_id IN 
      ('Subsumes', 'Is a');
  ",
  verbose = FALSE,
  render_sql = FALSE
)
```

```{r,echo=FALSE,results='markup'}
print_dt(taxonomy_relationship)
```




However, the CONCEPT_RELATIONSHIP table can only 
represent direct relationships that are 1 degree of separation 
between a concept. The CONCEPT_ANCESTOR table is used in situations 
where traversing the taxonomy beyond a single level is required.  



# Mapping 

The `Maps to` and `Mapped from` relationship can be used to 
either find the new concept assigned to a deprecated concept 
or the mapping between a non-standard vocabulary such as an 
ICD billing code to SNOMED. 

## Source Concepts to Target Concepts  

For 4115276, a _Mapped from_ 
relationship exists from MeSH, ICD-O-3, HemOnc, other SNOMED 
concepts, etc.... These relationships allows one to leverage 
existing codes in source systems to route them to the correct 
mapping in the OMOP CDM during an ETL.  

```sql
SELECT relationship_id, c.* 
FROM omop_vocabulary.concept_relationship cr 
LEFT JOIN omop_vocabulary.concept c 
ON c.concept_id = cr.concept_id_2
WHERE 
  cr.concept_id_1 = 4115276 AND
  cr.invalid_reason IS NULL AND 
  relationship_id IN 
    ('Mapped from', 'Maps to');
```

```{r}
df <- 
queryAthena(
  "
  SELECT relationship_id, c.* 
  FROM omop_vocabulary.concept_relationship cr 
  LEFT JOIN omop_vocabulary.concept c 
  ON c.concept_id = cr.concept_id_2
  WHERE 
    cr.concept_id_1 = 4115276 AND
    cr.invalid_reason IS NULL AND 
    relationship_id IN 
      ('Mapped from', 'Maps to');
  ",
  verbose = FALSE,
  render_sql = FALSE
)
```

```{r,echo=FALSE,results='markup'}
print_dt(df)
```




Another example of this type of relationship are mappings 
between NDC codes and RxNorm. Here, I find all the NDC concepts 
for _acetaminophen_.  

```sql
SELECT c.* 
FROM omop_vocabulary.concept_synonym cs 
INNER JOIN omop_vocabulary.concept c 
ON c.concept_id = cs.concept_id
WHERE 
  cs.concept_synonym_name LIKE '%acetaminophen%' 
    AND c.vocabulary_id = 'NDC' 
    AND c.invalid_reason IS NULL
;
```

```{r}
df <-
queryAthena(
  "
  SELECT c.* 
  FROM omop_vocabulary.concept_synonym cs 
  INNER JOIN omop_vocabulary.concept c 
  ON c.concept_id = cs.concept_id
  WHERE 
    cs.concept_synonym_name LIKE '%acetaminophen%' 
      AND c.vocabulary_id = 'NDC' 
      AND c.invalid_reason IS NULL
  ;
  ",
  verbose = FALSE,
  render_sql = FALSE
)
```


```{r,echo=FALSE,results='markup'}
print_dt(df)
```


The CONCEPT_RELATIONSHIP table is used to map NDC 
codes to RxNorm, which is often the case in an ETL from 
a source drug orders table to the OMOP DRUG_EXPOSURES table.  

```sql
SELECT 
  c1.concept_code AS ndc_code, 
  c1.concept_name AS ndc_drug,
  cr.relationship_id,
  c2.concept_code AS target_code,
  c2.concept_name AS target_name,
  c2.vocabulary_id AS target_vocabulary
FROM omop_vocabulary.concept_relationship cr 
LEFT JOIN omop_vocabulary.concept c1 
ON c1.concept_id = cr.concept_id_1  
LEFT JOIN omop_vocabulary.concept c2
ON c2.concept_id = cr.concept_id_2  
WHERE cr.concept_id_1 IN (
  SELECT DISTINCT c.concept_id  
  FROM omop_vocabulary.concept_synonym cs 
  INNER JOIN omop_vocabulary.concept c 
  ON c.concept_id = cs.concept_id
  WHERE 
    cs.concept_synonym_name LIKE '%acetaminophen%' 
      AND c.vocabulary_id = 'NDC' 
      AND c.invalid_reason IS NULL
  )
;
```

```{r}
df <- 
queryAthena(
  "
  SELECT 
    c1.concept_code AS ndc_code, 
    c1.concept_name AS ndc_drug,
    cr.relationship_id,
    c2.concept_code AS target_code,
    c2.concept_name AS target_name,
    c2.vocabulary_id AS target_vocabulary
  FROM omop_vocabulary.concept_relationship cr 
  LEFT JOIN omop_vocabulary.concept c1 
  ON c1.concept_id = cr.concept_id_1  
  LEFT JOIN omop_vocabulary.concept c2
  ON c2.concept_id = cr.concept_id_2  
  WHERE cr.concept_id_1 IN (
    SELECT DISTINCT c.concept_id  
    FROM omop_vocabulary.concept_synonym cs 
    INNER JOIN omop_vocabulary.concept c 
    ON c.concept_id = cs.concept_id
    WHERE 
      cs.concept_synonym_name LIKE '%acetaminophen%' 
        AND c.vocabulary_id = 'NDC' 
        AND c.invalid_reason IS NULL
    )
  ;
  ",
  verbose = FALSE,
  render_sql = FALSE
)
```

```{r,echo=FALSE,results='markup'}
print_dt(df)
```



## Deprecated Concepts to Valid Concepts 

In addition to the use case above, concepts that are deprecated 
usually have a _Maps to_ relationship to a valid replacement.  
Since a search for _NSCLC_ does not return any deprecated concepts, 
a search for _diabetes_ is used to exemplify instead.  

```sql
SELECT DISTINCT c.* 
FROM omop_vocabulary.concept_synonym cs  
INNER JOIN omop_vocabulary.concept c 
ON c.concept_id = cs.concept_id 
WHERE 
  cs.concept_synonym_name LIKE '%diabetes%' 
  AND c.invalid_reason IS NOT NULL;
```

```{r}
df <-
queryAthena(
  "
  SELECT DISTINCT c.* 
  FROM omop_vocabulary.concept_synonym cs  
  INNER JOIN omop_vocabulary.concept c 
  ON c.concept_id = cs.concept_id 
  WHERE 
    cs.concept_synonym_name LIKE '%diabetes%' 
    AND c.invalid_reason IS NOT NULL;
  ",
  verbose = FALSE,
  render_sql = FALSE
)
```

```{r,echo=FALSE,results='markup'}
print_dt(df)
```


For these deprecated _diabetes_ concepts, the _Maps to_ 
relationship exposes each of their valid counterparts.  

```sql
SELECT 
  c1.concept_code AS deprecated_concept_code, 
  c1.concept_name AS deprecated_concept_name,
  c1.vocabulary_id AS deprecated_vocabulary_id, 
  cr.relationship_id,
  c2.concept_code,
  c2.concept_name,
  c2.vocabulary_id,
  c2.invalid_reason
FROM omop_vocabulary.concept_relationship cr 
INNER JOIN omop_vocabulary.concept c1 
ON cr.concept_id_1 = c1.concept_id  
INNER JOIN omop_vocabulary.concept c2
ON cr.concept_id_2 = c2.concept_id  
WHERE 
    cr.concept_id_1 IN (
      SELECT DISTINCT c.concept_id 
      FROM omop_vocabulary.concept_synonym cs  
      INNER JOIN omop_vocabulary.concept c 
      ON c.concept_id = cs.concept_id 
      WHERE 
        cs.concept_synonym_name LIKE '%diabetes%' 
        AND c.invalid_reason IS NOT NULL
     )
       AND cr.invalid_reason IS NULL 
       AND cr.relationship_id = 'Maps to'
;
```

```{r}
df <-
queryAthena(
  "
  SELECT 
    c1.concept_code AS deprecated_concept_code, 
    c1.concept_name AS deprecated_concept_name,
    c1.vocabulary_id AS deprecated_vocabulary_id, 
    cr.relationship_id,
    c2.concept_code,
    c2.concept_name,
    c2.vocabulary_id,
    c2.invalid_reason
  FROM omop_vocabulary.concept_relationship cr 
  INNER JOIN omop_vocabulary.concept c1 
  ON cr.concept_id_1 = c1.concept_id  
  INNER JOIN omop_vocabulary.concept c2
  ON cr.concept_id_2 = c2.concept_id  
  WHERE 
      cr.concept_id_1 IN (
        SELECT DISTINCT c.concept_id 
        FROM omop_vocabulary.concept_synonym cs  
        INNER JOIN omop_vocabulary.concept c 
        ON c.concept_id = cs.concept_id 
        WHERE 
          cs.concept_synonym_name LIKE '%diabetes%' 
          AND c.invalid_reason IS NOT NULL
       )
         AND cr.invalid_reason IS NULL 
         AND cr.relationship_id = 'Maps to'
    
  ;
  ",
  verbose = FALSE,
  render_sql = FALSE
)
```

```{r,echo=FALSE,results='markup'}
print_dt(df)
```


# Lateral Relationships  

These relationships are all other relationships that are 
non-hierarchical, meaning that a corresponding entry 
would not be found in the CONCEPT_ANCESTOR table.  

```sql
SELECT relationship_id, c.* 
FROM omop_vocabulary.concept_relationship cr 
LEFT JOIN omop_vocabulary.concept c 
ON c.concept_id = cr.concept_id_2
WHERE 
  cr.concept_id_1 = 4115276 AND
  cr.invalid_reason IS NULL AND 
  relationship_id NOT IN 
    ('Subsumes', 'Is a', 'Mapped from', 'Maps to');
```


```{r}
df <-
queryAthena(
  "
  SELECT relationship_id, c.* 
  FROM omop_vocabulary.concept_relationship cr 
  LEFT JOIN omop_vocabulary.concept c 
  ON c.concept_id = cr.concept_id_2
  WHERE 
    cr.concept_id_1 = 4115276 AND
    cr.invalid_reason IS NULL AND 
    relationship_id NOT IN 
      ('Subsumes', 'Is a', 'Mapped from', 'Maps to');
  ",
  verbose = FALSE,
  render_sql = FALSE
)
```

```{r,echo=FALSE,results='markup'}
print_dt(df)
```

