---
title: "Mapping to HemOnc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping to HemOnc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(chariot)
```

```{r}
conn <- connectAthena()
```

Mapping Regimens to the HemOnc Ontology as represented in the OMOP Vocabularies 
can be approached several different ways. The `grepl` approach uses pattern 
matching in lieu of the ontological relationships to find the appropriate HemOnc 
counterpart to the input. 


```{r}
ex1 <- ho_grep_regimens("Trastuzumab",
                 conn = conn)
```

In this example, all Trastuzumab monotherapy regimens in HemOnc are searched for. 
The monotherapy is assumed by the length of the input (`1`), and the function 
first searches for "trastuzumab" and runs an additional search for the pattern
"monotherapy" based on the value of `1`. 

```{r}
ex1
```

If the search was intended to be for all regimens containing trastuzumab with a
known number of additional components, a `component_count` value can be provided. 
For example, if I want to know all the HemOnc regimens that contain 2 components, 
with 1 being trastuzumab: 

```{r}
ex2 <- 
  ho_grep_regimens("Trastuzumab",
                   conn = conn, 
                   component_count = 2)
```

The pattern matched is "Trastuzumab" with any HemOnc Regimens containing the 
pattern "and" for 2 components. 

```{r}
ex2
```

For searches for component counts greater than 2, the pattern matched is the 
number of commas. HemOnc Regimens composed of greater than 2 components are 
named by a comma-separated string.  

```{r}
ex3 <- 
  ho_grep_regimens("Trastuzumab",
                   conn = conn, 
                   component_count = 3)
```

For trastuzumab regimens containing 3 components, 2 commas are searched for. 

```{r}
ex3
```

To get all the possible combinations in regimens containing the component, 
`component_count` can be set to infinity. 

```{r}
ex4 <- 
  ho_grep_regimens("Trastuzumab",
                   conn = conn, 
                   component_count = Inf)
```

For all HemOnc functions, guardrails are put in place to prevent searches for 
components that may have been misspelled or subject to a typo. Here, I have 
mistyped trastuzumab. 

```{r,error=TRUE}
ho_grep_regimens("tastuzumab",
                 conn = conn)
```

You can also look up the Components that belong to a Regimen in HemOnc.  
```{r}
ex5 <- ho_lookup_antineoplastics(35804201,
                          conn = conn)
```

```{r}
ex5
```


The inverse can also be done, where a Regimen comprised by a set of Component 
Concept Ids is derived. 

```{r}
ex6 <-
ho_lookup_regimen(35803229,35803361,
                  conn = conn)
```

```{r}
ex6
```

These functions also allow Concept Class object support. For example, for the 
most recent example, using a Concept Class object may allow the user to stay 
informed with the Concept attributes when executing the function.  

```{r}
Paclitaxel <- get_concept(concept_id = 35803229)
Trastuzumab <- get_concept(concept_id = 35803361)
```

Now we know what is being looked up while also containing the `concept_id` needed 
as the essential input for the function.  

```{r}
Paclitaxel
```

```{r}
Trastuzumab 
```


```{r}
ex7 <-
ho_lookup_regimen(Paclitaxel,
                  Trastuzumab, 
                  conn = conn)
```

```{r}
ex7
```


```{r}
dcAthena(conn = conn)
```

