---
title: "Plot an ATC Classification"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"  
vignette: >
  %\VignetteIndexEntry{plot-atc-classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = "hide"
)
```

```{r setup}
library(chariot)
library(tidyverse)
conn <- chariot::connectAthena()
```

The target ATC Class first should be identified and a concept class object is derived. The Concept Id `21601386` for `ANTINEOPLASTIC AND IMMUNOMODULATING AGENTS` is used in this example.    

```{r concept_class_obj}
atc_class_concept <- get_concept(21601386,
                          vocab_schema = "omop_vocabulary",
                          conn = conn)
```

A `concept` class object contains slots for each field in the original Concept table with data type constraints. Concepts are handled as S4 classes in this package to allow the end-user to have access to all the concept attributes at any given point during their analysis. The concept class object of the example is as follows:

```{r concept_class_obj2}
atc_class_concept
```

The `atc_class_concept` is used as an argument for the `preview_atc_classification` function first to understand the number of levels that fall below this concept in the hierarchy, as well as the number of descendants at each of these levels to apply filters to create a manageable plot.  

```{r preview}
preview_atc_classification(concept_class_obj = atc_class_concept,
                     conn = conn)
```

I will now plot levels 1 to 5. 

```{r plot}
temp_html <- tempfile(fileext = ".html")
plot_atc_classification(conn = conn,
                          concept_class_obj = atc_class_concept,
                          range = 1:5,
                          file = temp_html)
```

```{r display}
htmltools::includeHTML(path = temp_html)
```



```{r final}
unlink(temp_html)
dcAthena(conn = conn)
```

