---
title: "Setup Drug Strength"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setup Drug Strength}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(chariot)
```

There are multiple transformations involved in using the Drug Strength 
table. Firstly, the conceptual representation of drug strength is spread across 
`amount`, `numerator`, and/or a `denominator`. Additionally, for each of the 3, 
there is a value and a unit, where the unit is represented as a concept id. 

The Drug Strength table is prepared in advance as the Drug Strength Staged table, 
where the `value` and `unit` are mapped and represented in their final form. 
For the staged `value`, the form takes on that of a varchar (string) with a mathematical 
expression of the fraction that can be evaluated later on.  

To stage the Drug Strength table, it is first processed. All the invalid entries 
in the Drug Strength table are filtered out and the related fields, `invalid_reason`, 
`valid_start_date`, and `valid_end_date` are removed. Since it has little relevance 
to the use cases seen in this package, `box_size` is also removed.  

The unit concept ids are also mapped to their concept names.

```{r,eval=FALSE}
ds_process()
```

The result of this function call is a Drug Strength Processed table. This is 
the table that is used to stage the `value` and `unit` fields and it is prerequisite 
for the next staging step. 

```{r,eval=FALSE}
ds_stage()
```

